\modLP{Identification of} Biological Regulatory Networks from Process Hitting models

### 2014-10-24 response.tex

\TODO{summary of response + changes}

Please find below our detailed answers.

\question{Beside, the definition of RT model, including signs that can be
deduced
from the K parameters must be justified or clarify. For a computer scientist
it is always surprising to have redundant pieces of information without accounting their matching.
With your definition I can write $a \xrightarrow+ b$ and I can
set $K_{b,\emptyset} = 1$ and $K_{b,\{a\}} = 0$.
On the one side I define a as an activator and
on the other side I define a as an inhibitor with K parameters. As signs do
not enter in the interpretation, why they are included in the formalism as
necessary ``comments''?}

\DISCUSS{LP}
Thomas' formalism with discrete parameters indeed includes some redundant piece of information in the sense that signs could be captured by the parameters with additional constraints, e.g. monotonous behaviors (which imposes an ordering between the value of parameters). Speaking from a computer science oriented point of view, the information encapsulated into a Thomas' network could be reduced to the set of nodes and the corresponding logical parameters. The single parametrization is very generic, ranging from totally decoupled specification to very constrained valuation w.r.t to the interaction graph. This would however be a disruption with the Thomas framework as it has been considered for more than 25 years.

Moreover signs and edges have been an inspiration for numerous works with regard to the analysis of networks. As showed for example in THOMAS, R. \& D'ARI, R., "Biological feedback", CRC Press, Boca Raton , Florida, USA, (1990), 316 pp), the analysis of some dynamical properties (e.g. sustained oscillations) can derive from the study of positive or negative circuits.

In addition, having both the interaction graph and the parametrization in the model allows to decouple the IG from the dynamics, which is interesting for modeling biological systems as the IG is often a prior knowledge with possible inconsistencies with the dynamics we want to encode with. A further step may then infer back the IG of the actual dynamics (here, an abstraction of the encoded dynamics) and compare it with the prior IG. Typically, if the inferred IG is included in the prior one, we have consistency.

Finally, several alternative definitions exist, with various degree of equivalency. As a consequence, we chose here to stick to the original formalism and picked this definition as it the original and standard one in the Thomas' community.

Nevertheless, to remove ambiguity and comply with the reviewer's suggestions, we redesigned section 2.2, emphasizing such a choice.

%\TODO{Answer inc. text modification}

@@

\answer
\DISCUSS{LP} yes, we have to mention those work, while stressing that they have
hard time with transient dynamics on very large networks, partly due to
concurrency, which makes model-checking very hard even with symbolic techniques.

\DISCUSS{LP} how such a discussion is relevant for the paper? Can't we simply
say that PH allows to model indeterministic Boolean and multi-valued networks
and remove (or drastically reduce) this discussion on model-checking?

\question{
(P.3 and 6) The term “cooperation” is often mentioned. It is not self-
explanatory for the reader. In particular in P.3 “associated influences” is
not explanatory at all. If I have missed it I suggest to more clearly emphasize it. This comment does not concern cooperative sort which is clearly
explained, but the notion of cooperation that cooperative sort encompasses.
A definition of “cooperation” is required (formal or not) or more explanations of the notion of cooperation in this context.
}

\DISCUSS{LP} we should avoid this term in introduction, and only use once
formally defined.
In the introduction, maybe we should mention that
PH models can embed additional automata that encode for some (partial) logic
gates. Such additional automata that can be considered as ``internal''
variables, and we may want to mask them in abstract representations of the
systems, such as IG and Thomas' parameters.
\DISCUSS{LP} the general the notion of cooperation should be defined in the
introduction.

### 2014-10-25 parts-infer-param.tex
some typos in K inference

\begin{property}[Well-formed PH for parameter inference]\label{pro:wf-ph-K}
A PH is well-formed for parameter inference if and only if
it is well-formed for BRN identification (\pref{pro:wf-ph}) and
that the associated IG $(\Gamma, E)$ verifies that:
\modLP{
\begin{align}
\begin{split}
\forall b\rightarrow a&\in E, b\neq a,
	\forall \sigma\in L(\PHpredecgene{a}),\\
    \forall (i,j & \in \levels{b}{a} \vee i,j \in \ulevels{b}{a}),
	\forall a_k \in L_a,
	\\
&	\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_i\})))
	=\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_j\})))
\end{split}
\\
\begin{split}
\forall a\in\Gamma, &\forall b\in\PHpredecgene{a},
b\neq a \wedge b\rightarrow a\notin E,\\
	\forall \sigma&\in L(\PHpredecgene{a}), 
    \forall b_i,b_j \in L_b,
	\forall a_k \in L_a,
	\\
&	\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_i\})))
	=\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_j\})))
\end{split}
\end{align}
}
\end{property}

@@

\modLP{The configuration of the PH corresponding to $\omega$ is given as follows.
For each component $b \in \GRNreg{a}$, we define $\sigma^b_{a,\omega}$ (\pref{eq:param_context}) as 
the process of $b$ with the level in 
$\levels{b}{a}$ if $b\in\omega$, or in $\ulevels{b}{a}$ if $b\not\omega$.
Because of \pref{pro:wf-ph-K}, if several possible levels exist, this process can be chosen
arbitrarily.
If $b$ does not regulate $a$ in the IG, all processes of sort $b$ should have the same action on
$a$, so the process $b_0$ is arbitrarily selected (\pref{eq:param_context_free}).
The configuration $\sigma_{a,\omega}$ corresponding to $\omega$ (\pref{eq:K-ctx})
is then obtained by extending the configuration of the
regulators of $a$ to the (deterministic) cooperative sorts (\pref{def:ctx-sigma} in
\pref{ssec:split-sorts}).
Finally, we denote by $S_a(\sigma)$ (\pref{eq:K-domain}) the set of processes of sort $a$ that are
compatible with a configuration $\sigma$: if $a$ is specified in $\ctx$, then $S_a(\sigma) = \{a_i\}$
where $a_i\in\sigma$; otherwise $S_a(\sigma)$ is the all the processes of sort $a$, $L_a$.}
\begin{align}
  \label{eq:param_context}
  \forall b\rightarrow a\in E,~
  \sigma_{a,\omega}^b & \DEF \begin{cases}
    \modLP{\min}(\levels{b}{a}) & \text{if $b \in \omega$,}\\
    \modLP{\min}(\ulevels{b}{a}) & \text{if $b \notin \omega$}
  \end{cases}
  \\
  \modLP{
  \forall b\in\PHpredecgene a, b\rightarrow a\notin E,~
  \sigma_{a,\omega}^b} & \modLP{\DEF b_0}
  \label{eq:param_context_free}
  \\
  \sigma_{a,\omega} & \DEF
  \modLP{\ctx_a^{\PHpredecgene{a}}(
  	\state{\sigma^b_{a,\omega} \mid  b\in\PHpredecgene{a}})}
  \label{eq:K-ctx}
  \\
  \modLP{
  \forall a\in\Gamma,~
  S_a(\sigma)} &\modLP{\DEF
	\begin{cases}
	\{a_i\} & \text{if }a_i\in \ctx\\
	L_a & \text{otherwise.}
	\end{cases}}
\label{eq:K-domain}
\end{align}

>>>>

\begin{property}[Well-formed PH for parameter inference]\label{pro:wf-ph-K}
A PH is well-formed for parameter inference if and only if
it is well-formed for BRN identification (\pref{pro:wf-ph}) and
that the associated IG $(\Gamma, E)$ verifies that:
\modLP{
\begin{align}
\begin{split}
\forall b\rightarrow a&\in E, b\neq a,
	\forall \sigma\in L(\PHpredecgene{a}),\\
    \forall (i,j & \in \levels{b}{a} \vee i,j \in \ulevels{b}{a}),
	\forall a_k \in L_a,
	\\
&	\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_i\}))
	=\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_j\}))
\end{split}
\\
\begin{split}
\forall a\in\Gamma, &\forall b\in\PHpredecgene{a},
b\neq a \wedge b\rightarrow a\notin E,\\
	\forall \sigma&\in L(\PHpredecgene{a}), 
    \forall b_i,b_j \in L_b,
	\forall a_k \in L_a,
	\\
&	\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_i\}))
	=\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_j\}))
\end{split}
\end{align}
}
\end{property}

@@

\modLP{The configuration of the PH corresponding to $\omega$ is given as follows.
For each component $b \in \GRNreg{a}$, we define $\sigma^b_{a,\omega}$ (\pref{eq:param_context}) as 
the process of $b$ with the level in 
$\levels{b}{a}$ if $b\in\omega$, or in $\ulevels{b}{a}$ if $b\notin\omega$.
Because of \pref{pro:wf-ph-K}, if several possible levels exist, this process can be chosen
arbitrarily.
If $b$ does not regulate $a$ in the IG, all processes of sort $b$ should have the same action on
$a$, so the process $b_0$ is arbitrarily selected (\pref{eq:param_context_free}).
The configuration $\sigma_{a,\omega}$ corresponding to $\omega$ (\pref{eq:K-ctx})
is then obtained by extending the configuration of the
regulators of $a$ to the (deterministic) cooperative sorts (\pref{def:ctx-sigma} in
\pref{ssec:split-sorts}).
Finally, we denote by $S_a(\sigma)$ (\pref{eq:K-domain}) the set of processes of sort $a$ that are
compatible with a configuration $\sigma$: if $a$ is specified in $\ctx$, then $S_a(\sigma) = \{a_i\}$
where $a_i\in\sigma$; otherwise $S_a(\sigma)$ contains all the processes of sort $a$, $L_a$.}
\begin{align}
  \label{eq:param_context}
  \forall b\rightarrow a\in E,~
  \sigma_{a,\omega}^b & \DEF \begin{cases}
    \modLP{\min}(\levels{b}{a}) & \text{if $b \in \omega$,}\\
    \modLP{\min}(\ulevels{b}{a}) & \text{if $b \notin \omega$}
  \end{cases}
  \\
  \modLP{
  \forall b\in\PHpredecgene a, b\rightarrow a\notin E,~
  \sigma_{a,\omega}^b} & \modLP{\DEF b_0}
  \label{eq:param_context_free}
  \\
  \sigma_{a,\omega} & \DEF
  \modLP{\ctx_a^{\PHpredecgene{a}}(
  	\state{\sigma^b_{a,\omega} \mid  b\in\PHpredecgene{a}})}
  \label{eq:K-ctx}
  \\
  \modLP{
  \forall a\in\Gamma,~
  S_a(\sigma)} &\modLP{\DEF
	\begin{cases}
	\{a_i\} & \text{if }a_i\in \ctx\\
	L_a & \text{otherwise.}
	\end{cases}}
\label{eq:K-domain}
\end{align}
